White Wine Quality Analysis by Mehdi Jafari 
======================================================== 
 
```{r echo=FALSE, message=FALSE, warning=FALSE, packages} 
library(ggplot2)  
library(gridExtra)  
library(knitr) 
library(dplyr) 
library(RColorBrewer)
library(reshape2) 
library(GGally) 
library(corrplot)
``` 
 
```{r echo=FALSE, message=FALSE, warning=FALSE} 
# Load the Data 
white_wine <- read.csv('wineQualityWhites.csv') 
``` 
 
This data set contains 4898 white wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine,  
providing a rating between 0 (very bad) and 10 (very excellent). 
 
# Univariate Plots Section 
 
Our dataset consists of eleven variables with 4898 observations. here is the  
list of all the variables. 
 
```{r echo=FALSE, message=FALSE, warning=FALSE} 
# exploring the variables

str(white_wine)

``` 


```{r echo=FALSE, message=FALSE, warning=FALSE} 
#dataset summary

summary(white_wine) 

``` 

 

###All Variables distribution   
 

Based on the following histograms, all variables have normal distribution except
Residual sugar which is right skewed distribution. In the dataset all white 
wins' quality varies between 3 to 9 with most of them have quality 6. In 
addition almost all variables have outliers (outliers are flagged with red in 
box plot) except alcohol.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=30, fig.width=9} 

plot_list <- list();
counter <- 1

for(col_name in colnames(white_wine)){
  
  if(col_name != 'X')
  {
    plot_list[[counter]] <- ggplot( aes_string( x = col_name),
                                    data = white_wine) +
      geom_histogram(col="red", fill="green")
    
    counter <- counter + 1
    
    plot_list[[counter]] <- ggplot( aes_string( x = col_name, y = col_name),
                                    data = white_wine) +
      geom_boxplot(outlier.colour="red", outlier.size=4)
    
    counter <- counter + 1
  }
}

grid.arrange(grobs = plot_list, ncol=2) 


 
``` 

# Univariate Analysis

### What is the structure of your dataset? 

our dataset contains 11 variables that you can find their short description in 
the following:

+ **fixed acidity**: most acids involved with wine or fixed or nonvolatile (do 
not evaporate readily). The distribution of fixed acidity is normal although
there are some outliers. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$fixed.acidity)
```

+ **volatile acidity**: the amount of acetic acid in wine, which at too high of
levels can lead to an unpleasant, vinegar taste. This variable seems to have 
many outliers as the mean is 0.2782 and the maximum value in the distribution 
is 1.1. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$volatile.acidity)
```

+ **citric acid**: found in small quantities, citric acid can add 'freshness' 
and flavor to wines. As we can see 75 percent of observations have less than 
0.3900 mg citric.acid while the max is 1.66 mg.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$citric.acid)
```

+ **residual sugar**: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater than
45 grams/liter are considered sweet. Based on the definition, none of the 
wines in our dataset is considered as sweet wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$residual.sugar)
```


+ **chlorides**: the amount of salt in the wine

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$chlorides)
```

+ **free sulfur dioxide**: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$free.sulfur.dioxide)
```

+ **total sulfur dioxide**: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$total.sulfur.dioxide)
```

+ **density**: the density of water is close to that of water depending on the 
percent alcohol and sugar content. this variable has only 3 outliers in our
dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$density)
```

+ **pH**: describes how acidic or basic a wine is on a scale from 0 (very 
acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$pH)
```

+ **sulphates**: a wine additive which can contribute to sulfur dioxide gas 
(S02) levels, which acts as an antimicrobial and antioxidant


```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$sulphates)
```

+ **alcohol**: the percent alcohol content of the wine. There is no outlier in 
alcohol distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine$alcohol)
```

### What is/are the main feature(s) of interest in your dataset? 

The main feature in our dataset is quality. In this report will try to find 
a correlation  between other variables and quality.

### What other features in the dataset do you think will help support your \ 
investigation into your feature(s) of interest? 

residual sugar, volatile acidity, density and alcohol could have correlation 
with quality of wine.

### Did you create any new variables from existing variables in the dataset? 
In order to have a better diagram, quality is categorized into four factors
 (bad (0-4), average (4-6), good (7-8) and excellent (9-10))

   
```{r  echo=FALSE, message=FALSE, warning=FALSE}

white_wine$quality.bucket <- cut(white_wine$quality,  

                       breaks = c(0, 4, 6, 8, 10),  

                       labels = c('bad','average','good','excellent')) 

```
### Of the features you investigated, were there any unusual distributions? \ 

Did you perform any operations on the data to tidy, adjust, or change the form \ 

of the data? If so, why did you do this? 

As it is mentioned earlier residual.sugar has right skewed distribution then a
log10 transformation is applied to this variable. the plot is depicted in the 
following:

```{r  echo=FALSE, message=FALSE, warning=FALSE}

p1 <- ggplot( aes( x = residual.sugar), data = white_wine) + 
  geom_histogram() 
 
p2 <- ggplot( aes( x = residual.sugar), data = white_wine ) + 
  geom_histogram()+
  scale_x_log10()


grid.arrange(p1,p2 ,ncol = 2)

```

# Bivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=9} 
ggcorr(white_wine[, 2:13], label = TRUE, label_size = 3, hjust = 0.8)
``` 
```{r  echo=FALSE, message=FALSE, warning=FALSE}

d_cor <- as.matrix(cor(white_wine[c(2, 3, 4,5,6,7,8,9,10,11,12)]))
d_cor_melt <- arrange(melt(d_cor), -abs(value))

dplyr::filter(d_cor_melt, value > .5) # .5 is arbitrary

# If you don't want to download dplyr
subset(d_cor_melt, value > .5 & value < 1)

```

As it is shown in the above mentioned result, the following pairs have the most 
correlation with each other:

+ density and residual.sugar
+ total.sulfur.dioxide and free.sulfur.dioxide
+ density and total.sulfur.dioxide

since free.sulfur.dioxide is a subset of total.sulfur.dioxide, we will not 
investigate those pairs.

```{r  echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = density, y = residual.sugar), 

       data = white_wine) +  

  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color',  

                            override.aes = list(alpha = 1, size = 2))) + 

  geom_jitter(alpha = 0.05, size = 2) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density, y = total.sulfur.dioxide), 

       data = white_wine) +  

  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color',  

                            override.aes = list(alpha = 1, size = 2))) + 

  geom_jitter(alpha = 0.1, size = 2)

```
Since the alcohol has the greatest correlation with quality which is the main
factor, plot the correlation of these two variable could be helpful.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(factor(quality), 
            alcohol), 
        data = white_wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```


# Bivariate Analysis 

### Talk about some of the relationships you observed in this part of the \ 

investigation. How did the feature(s) of interest vary with other features in \ 

the dataset? 

Based on the result we may say that alcohol has the most correlation with the 
quality (0.4355747) and has the least correlation with citiric.acid
(0.009209091).
 

### Did you observe any interesting relationships between the other features \ 

(not the main feature(s) of interest)? 

As it is represented in the two previous plots, the more residual.sugar and 
total.sulfur.dioxide, the more density.
 

### What was the strongest relationship you found? 
As it was mentioned the strongest relationship is between density and 
residual.sugar which is 0.8389665

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = residual.sugar, y = volatile.acidity, color = quality.bucket),
       data = white_wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0,25)


```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density, y = alcohol, color = quality.bucket),
       data = white_wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0.98,1.01)

```
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = residual.sugar, y = alcohol, color = quality.bucket),
       data = white_wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0,25)

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = density, color = quality.bucket),
       data = white_wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            )) +
  geom_jitter(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0, 25) +
  ylim(0.98 , 1.01)

```

# Multivariate Analysis

### Relationships observed in investigation

As it was represented in the plots we can see the good white wins have less 
than 20mg of residual.sugar and less than 7.5 volatile acidity. Also, Good 
wines tend to have more alcohol and less density.

### Interesting or surprising interactions between features?

As it was mentioned earlier, the combination of alcohol, density and alcohol 
affects the quality of wine.


# Final Plots and Summary

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE}


ggplot(aes(factor(quality), 
            alcohol), 
        data = white_wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab('Wine Quality') +
  ylab('Alcohol Percentage') +
  labs(title = 'Alcohol Percentage by Quality')

```

### Description One

This diagram depicts the correlation between quality and alcohol. Excellent wines
seem to have between 11 and 14 percent alcohol.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = residual.sugar, y = alcohol, color = quality.bucket),
       data = white_wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0,25) +
  xlab('Residual Suger') +
  ylab('Alcohol Percentage') +
  labs(title = 'Residual Suger and Alcohol Percentage by Quality Categories')

```

### Description Two
High quality wines have less than 20mg residual.suger. As it is presented in 
the diagram the less residual.suger and the more alcohol the more chance you 
have a better quality wine.


### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = residual.sugar, y = volatile.acidity, color = quality.bucket),
       data = white_wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0,25) +
  xlab('volatile acidity (acetic acid - g / dm^3)') +
  ylab('Residual Suger') +
  labs(title = 'Residual Suger and volatile acidity (acetic acid - g / dm^3) by Quality Categories')

```


### Description Three
High quality wines have less than 8g volatile acidity. As it is presented in the
diagram the less residual.sugar and the less volatile acidity the more chance 
you have a better quality wine.


# Reflection


In this project, I analyzed the effect of different variables of white wine on 
its quality based on the dataset that contained 4,898 observations with 11 
variables on the chemical properties. I categorized the quality of wine into 4 
groups (bad, average, good, excellent). I found out that good wines tend to have
more alcohol and less sugar.

I can't expand my findings to the other white wine datasets since my sample is
not a random dataset from all type of white wines, it's a sample of wines of a 
specific region in Portuguese called Vinho Verde. And, there might be a 
combination of several variables (more than two) would outcome to a perfect that
wasn't investigated in this project.
