Red Wine Quality Analysis by Mehdi Jafari 
======================================================== 
 
```{r echo=FALSE, message=FALSE, warning=FALSE, packages} 
library(ggplot2)  
library(gridExtra)  
library(knitr) 
library(dplyr) 
library(RColorBrewer)
library(reshape2) 
library(GGally) 

``` 
 
```{r echo=FALSE, message=FALSE, warning=FALSE} 
# Load the Data 
white_wine <- read.csv('wineQualityWhites.csv') 
``` 
 
This data set contains 4898 white wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine,  
providing a rating between 0 (very bad) and 10 (very excellent). 
 
# Univariate Plots Section 
 
Our dataset consists of threen variables with 4898 observations. here is the  
list of all the variables. 
 
```{r echo=FALSE, message=FALSE, warning=FALSE} 
# exploring the variables 
str(white_wine) 

names(white_wine) 
``` 
 

```{r echo=FALSE, message=FALSE, warning=FALSE} 

summary(white_wine) 
``` 
 
###All Variables distribution   
 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=30, fig.width=9} 

plot_list <- list();
counter <- 1

for(col_name in colnames(white_wine)){
  
  if(col_name != 'X')
  {
    plot_list[[p]] <- ggplot( aes_string( x = col_name), data = white_wine) + 
      geom_histogram(col="red", fill="green")
    counter <- counter + 1
    
    plot_list[[counter]] <- ggplot( aes_string( x = col_name, y = col_name), data = white_wine) +
      geom_boxplot(outlier.colour="red", outlier.size=4)
    counter <- counter + 1
  }
}

grid.arrange(grobs = plot_list, ncol=2) 


 
``` 
 